\bf Teaching analysis reproducibility with accessibility: Enriching secondary documentation to support skill level diversity of users.

### 2022-02-10 jsdse/Manuscript.tex
figure: tree from tutorial

\bigskip
\begin{abstract}
Research reproducibility is essential for scientific development. Yet, rates of reproducibility
 are low, especially in the natural sciences. In the current scientific landscape,
most research relies on computing tools and software. Hence, efforts for increasing
reproducibility rates have focused on providing availability of scientific workflows
as computer code, as well as raw and processed data in a digital form.
However, making these research products widely available, usually online, does not
necessarily ensures that
they can be easily reached by the majority of the population,
making them effectively inaccessible, counteracting availability and hindering reproducibility.

In this paper, we argue that accessibility of resources is a key aspect of research reproducibility,
yet, to our knowledge, it has been widely overlooked.
We recognize that accessibility has many angles (social, psychological,
education level), but we focus on lack of accessibility at the skill and pedagogical level.
Using an example within evolutionary biology, we identify barriers to
accessibility in the natural sciences:
%% list the barriers
highly specialized jargon,
Then, we propose a set of principles to improve accessibility of code and data,
and illustrate their application in a set of teaching materials for evolutionary biology that we put
to work in the field.
Finally, we elaborate on the general application of these principles for
documenting scientific workflows, to provide present learners
and future researchers with tools for successful scientific reproducibility.
%> ** add a sentence on implications for teaching reproducibility

@@

Reproducibility --the extent to which consistent results are obtained when a scientific
research experiment or workflow is repeated \citep{repdef2021}--
is a key aspect of the advancement
of science, as it constitutes a minimum standard that allows understanding research products
(e.g., methods, data, analysis, results, etc.)
to determine their reliability and generality, and eventually build up scientific
knowledge and applications based on those products
\citep{king1995replication, peng2011reproducible, powers2019open}.
In the natural sciences, rates of reproducibility are low \citep{ioannidis2005most, prinz2011believe},
which has elicited concerns about a reproducibility crisis in the field \citep{baker2016reproducibility}.
%%% The scientific community has united to incentivize cultural changes that will improve
%%% reproducibility rates long term, such as transparency, availability, and workflow
%%% automatization, to name a few \citep{peng2015reproducibility}.
In response, the scientific community has been developing new principles and standards to incentivize
cultural changes in an effort to improve long term reproducibility rates in the natural sciences
\citep{peng2015reproducibility, wilkinson2016fair}.
A standard for reproducibility that has received much attention is scientific availability, which
we define as a property denoting that a research product can be acquired, copied, analyzed,
processed and/or reused, at no financial, legal or technical cost \citep{arnold2019turing},
and with no geographic, demographic or temporal barriers for the population \citep{fecher2014open}.
In this paper, we argue that availability can not be fully achieved without accessibility (Box 1).
We identify factors that have particularly affected accessibility in
the natural sciences, and ways the scientific community can address them to create reproducible
scientific workflows. These principles can be generalized and integrated
into the undergraduate and graduate school STEM curriculum, either for courses specialized
in reproducibility or within other subject areas as a necessary component of
successful and impactful science.

@@

To elaborate on our thesis,
%% we use a common scientific workflow in phylogenetics,
%% ** Jessica mentions that workflow is confusing as she thought it referred to the OpenTree
%% workflow, but we actually mean the more general workflow of reconstructing a phylogenetic tree.
%% She suggests something more general:
we choose an example from the research field of phylogenetics,
a discipline within evolutionary biology. Phylogenetics focuses
on investigating the history of shared ancestry of living organisms from biological data,
and represents
this evolutionary history in a diagram known as phylogeny (a figure from the tutorial).
Phylogenies provide the
basis to study and understand all biological processes in an evolutionary context
\citep{dobzhansky1973nothing}.
Given the importance of phylogenies for biological research, improving reproducibility rates
in phylogenetics is increasingly relevant for many aspects of the natural sciences.

@@

The Open Tree of Life project (OpenTree) has developed a computational platform that provides
availability of results from vetted, state-of-the-art research in phylogenetics
\citep{opentreeoflife2019synth}.
%% by creating a standardized database of phylogenetic data that is used to synthesize
%% a single phylogenetic tree encompassing all life
OpenTree's database, the Phylesystem \citep{mctavish2015phylesystem}, provides
accessibility of phylogenetic data by standardizing organism names across biological databases,
and supporting digital data obtention in a computer readable format.
While this is accessible through their Graphical User Interface,
It does not support reproducibility
Accessibility does not imply reproducibility
In order for something to be highly reproducible, it has to be largely accessible.

@@

@@

A story-like narrative that links pieces of code together and invites learners
to try the code can lead learners to remember what they are doing and
why they are doing it.
This ``literate programming'' paradigm \citep{knuth1984literate, fritzson2002mathmodelica}
makes code more approachable, as it integrates narratives with computer code in
the same document, supporting learners in actively following, remembering and understanding
the code usage \citep{piccolo2016tools}.
Documents developed with ``literate programming'' can be made more accessible by
choosing narratives that are relatable to a more general audience.

We examined available primary documentation for the package \texttt{rotl},
and designed a narrative that required the usage of as many functions as possible.
We demonstrate code applications that are commonly requested by OpenTree users,
but that are not demonstrated in the R package primary documentation.
By framing the function workflow using highly requested uses, the documentation acquires a
narrative arc that is easier to follow and remember by users. This can also facilitate the application
of code to other use cases in biology of interest for the learners.

%% Maybe Figure!! For the tutorial demonstrated here, we used the commonly requested use case of obtaining
%% a phylogenetic tree for all lineages within a specific taxonomic rank.

@@

A practice that has become more and more widespread in programming languages pedagogical practices
is the use of typos and mistakes to normalize them for learners and show them how
to solve them when they are outside the classroom \citep{shannon2015live}.
Yet, this is rarely done for written pedagogical materials.
Primary documentation focuses on demonstrating usage function with examples that
work seamlessly, without errors. We argue that the opposite is needed to support
adoption of reproducible workflows and support long term independence in learners
\citep{gaspar2007restoring}.
We demonstrate examples that do not work
as expected and exemplify ways to address them (Figure \ref{fig:first}).

>>>>

\bigskip
\begin{abstract}
Research reproducibility is essential for scientific development. Yet, rates of reproducibility
are low, especially in the natural sciences. As increasingly more research is relying on
computing tools and software, efforts for improving
reproducibility rates have focused on providing availability of research workflows
as computer code, as well as raw and processed data in a digital form.
However, making research products digitally available, does not
necessarily ensures that
they can be easily reached by the majority of the target user base, let alone the general population,
making them effectively inaccessible, counteracting availability, and hindering reproducibility.

In this paper, we argue that accessibility of resources is a key aspect for
successful research reproducibility.
While accessibility is a broad term that encompasses various aspects
of the human experience (visual, motor, auditory, cognitive, social, psychological,
educational, etc.), we particularly address barriers to accessibility at the
technological, social and pedagogical level.
Using an example within evolutionary biology, we identify barriers to
accessibility in the natural sciences:
%% list the barriers
specialized jargon, user programming experience, cognitive load,

Then, we propose a set of principles to improve accessibility of code and data,
and illustrate their application in a set of teaching materials for evolutionary biology that we put
to work in the field.
Finally, we elaborate on the general application of these principles for
documenting scientific workflows, to provide present learners
and future researchers with tools for successful scientific reproducibility.
%> ** add a sentence on implications for teaching reproducibility

@@

Research reproducibility --the extent to which consistent results are obtained when
a scientific experiment or research workflow is repeated \citep{repdef2021}--
is a key aspect of the advancement
of science, as it constitutes a minimum standard that allows understanding research products
(e.g., methods, data, analysis, results, etc.)
to determine their reliability and generality, and eventually build up scientific
knowledge and applications based on those products
\citep{king1995replication, peng2011reproducible, powers2019open}.
In the natural sciences, rates of reproducibility are low \citep{ioannidis2005most, prinz2011believe},
which has elicited concerns about a crisis in the field \citep{baker2016reproducibility}.
%%% The scientific community has united to incentivize cultural changes that will improve
%%% reproducibility rates long term, such as transparency, availability, and workflow
%%% automatization, to name a few \citep{peng2015reproducibility}.
In response, the scientific community has been developing new principles and standards to incentivize
cultural changes in an effort to improve long term reproducibility rates in the natural sciences
\citep{peng2015reproducibility, wilkinson2016fair}.
A standard for reproducibility that has received much attention is scientific availability, which
we define as a property denoting that a research product can be reached (acquired, copied, analyzed,
processed and/or reused) at no financial, legal or technical cost \citep{arnold2019turing},
and without geographic, demographic, social or temporal barriers for the population \citep{fecher2014open}.
In this paper, we argue that availability can not be fully achieved without accessibility (Box 1).

We recognize that accessibility is a term that encompasses various aspects
of the human experience (visual, motor, auditory, cognitive, social, psychological,
educational, etc.).
However, some forms of accessibility do not support reproducibility. For example,
Graphical User Interfaces
and
we focus on X types of accessibility

% not all types of accessibility support reproducibility
% while accessibility is key to support availability, and availability is ket for reproducibility,
% some forms of accessibility do not support reproducibility.

We identify factors that have particularly affected accessibility in
the natural sciences, and ways the scientific community can address them to create reproducible
research workflows. These principles can be generalized and integrated
into the undergraduate and graduate school STEM curriculum, either for courses specialized
in reproducibility or within other subject areas, as a necessary component of
successful and impactful science.

@@

\section*{An example from phylogenetics}
\label{sec:example}

To elaborate on our thesis,
%% we use a common scientific workflow in phylogenetics,
%% ** Jessica mentions that workflow is confusing as she thought it referred to the OpenTree
%% workflow, but we actually mean the more general workflow of reconstructing a phylogenetic tree.
%% She suggests something more general:
we choose an example from the research field of phylogenetics,
a discipline within evolutionary biology. Phylogenetics focuses
on investigating the history of shared ancestry of living organisms from biological data,
and represents
this evolutionary history in a diagram known as phylogeny (Figure \ref{fig:tree}).
Phylogenies provide the
basis to study and understand all biological processes in an evolutionary context
\citep{dobzhansky1973nothing}. Hence,
% Given the importance of phylogenies for biological research,
it appears that improving reproducibility rates
in phylogenetics has the potential to positively impact research across the natural sciences.

@@

The Open Tree of Life project (OpenTree) has developed a computational platform that provides
availability of results from vetted, state-of-the-art research in phylogenetics
\citep{opentreeoflife2019synth}.
%% by creating a standardized database of phylogenetic data that is used to synthesize
%% a single phylogenetic tree encompassing all life
OpenTree's database, the Phylesystem \citep{mctavish2015phylesystem}, provides
accessibility of phylogenetic data by standardizing organism names across biological databases,
and supporting digital data obtention in a computer readable format.

While this is accessible through their Graphical User Interface,
It does not support reproducibility

Accessibility does not imply reproducibility

In order for something to be highly reproducible, it has to be largely accessible.

@@

\begin{figure}
\begin{center}
\includegraphics[width=3in]{fig-tree.png}
\end{center}
\caption{A phylogenetic tree from our tutorial website, extracted using OpenTree of Life tools and the rotl R package. \label{fig:second}}
\end{figure}

@@

A story-like narrative that links pieces of code together and invites learners
to try the code, can lead learners to remember what they are doing and
why they are doing it.
This ``literate programming'' paradigm \citep{knuth1984literate, fritzson2002mathmodelica}
makes code more approachable, as it integrates narratives with computer code in
the same document, supporting learners in actively following
the code usage, supporting memory and understanding \citep{piccolo2016tools}.

We propose that documents developed with ``literate programming'' can be made more
accessible by choosing narratives that are relatable to a more general audience.
An easy way to do this in biology is choosing as model organism a taxon that a
research group is more interested in studying. For the non-specialized user base,
a highly charismatic group such as dinosaurs will do the trick.

We examined available primary documentation for the package \texttt{rotl},
and designed a narrative that required the usage of as many functions as possible.
We demonstrate code applications that are commonly requested by OpenTree users,
but that are not demonstrated in the primary documentation of the R package.
By framing the function workflow using highly requested uses, the documentation acquires a
narrative arc that is easier to follow and remember by users. This can also facilitate
the application of code to other use cases in biology of interest for the users.

%% Maybe Figure!! For the tutorial demonstrated here, we used the commonly requested
%% use case of obtaining
%% a phylogenetic tree for all lineages within a specific taxonomic rank.

@@

A practice that has become more and more widespread in programming-language pedagogical practices
is the use of typos and mistakes to normalize them for learners, and show them how
to solve them when they are outside the classroom \citep{shannon2015live}.
Yet, this is rarely done on written pedagogical materials.
Primary documentation focuses on demonstrating usage function with examples that
work seamlessly, without errors. We argue that the opposite is needed to support
adoption of reproducible workflows and support long term independence in learners
\citep{gaspar2007restoring}.
We demonstrate examples that do not work
as expected and exemplify ways to address them (Figure \ref{fig:first}).
